---
title: "nlcd_oba"
author: "egb"
date: "2024-11-20"
output:
  pdf_document: default
  html_document: default
---
```{r load-libraries, echo=FALSE, results="hide", message=FALSE, warning=FALSE}
library(terra)
library(ggplot2)
library(dplyr)
library(sf)
library(tidyr)
library(vegan)
library(tibble) 

```

# Load in Oregon basemap
```{r plot-or}
or_boundary <- read_sf(dsn = "data/Oregon_State_Boundary_-2975945127261102476/Oregon_State_Boundary.shp")
# str(or_boundary)

```

# Load in OBA Data
```{r oba-import}
oba <-
  read.csv("data/OBA_2018-2023_combined_dataset_working.csv")
```

# Segment OBA data
```{r oba-segment}

# - Metadata we need:
#   - `Species`: The taxonomic identification of each bee (group of closely related species)
#   - `Genus`: The taxonomic identification of each bee (specific type within Genus)
#   - `Elevation`: The elevation (in meters) at which each bee was collected.
#   - `Coordinates`: Latitude (`Dec.Lat.`) and longitude (`Dec.Long.`) in decimal degrees.

# Subset oba data to our needed metadata
oba <- dplyr::select(oba, Genus, Species, Dec..Lat., Dec..Long., Elevation..m., MonthJul)
colnames(oba) <- c("Genus", "Spec", "Lat", "Long", "Elev", "Month")

# Make elevation values numeric 
oba$Elev <- as.numeric(oba$Elev)

```

# Clean OBA data
```{r oba-clean}
# Remove elevation rows with null values 

for (col in colnames(oba)) {
  oba <- oba[oba[[col]] != "", ]
  oba <- oba[!is.na(oba[[col]]), ]
}
# Number of rows after cleaning: 41643
# nrow(oba)
# head(oba, 10)
```

# Combine genus & species columns in OBA data
```{r genus-spec}
# Combine genus & species 
GenusSpec <- paste(oba$Genus, oba$Spec, sep=" ")

oba <- oba %>%
  mutate(GenusSpec)

# colnames(oba)
#nhead(oba, 10)
```

# Load in NLCD - Oregon (National Land Cover Database)
```{r nlcd-import, include=FALSE}
library(raster)

# Observe the dataset metadata here: https://www.arcgis.com/sharing/rest/content/items/9bbaa64718774bbfbf5c6ade0edf86d3/info/metadata/metadata.xml?format=default&output=html 

# Load landcover data as raster 
nlcd <- rast("data/NLCD_2016_Land_Cover_OR/NLCD_2016_Land_Cover_OR.img")

# Check original resolution
# Original: 98.4252 98.4252 ft
res(nlcd)
# cat("Original resolution:", original_res, "\n")


```

# NLCD Manipulation
```{r}
# Create a new template raster with the desired resolution
# The larger the resolution, the easier to load

# NOTE: need to reproject the raster before changing the crs to use meters as the unit 
target_res <- 500  # Desired resolution in original map units (feet)
extent_nlcd <- ext(nlcd)  # Get the extent of the original raster

# Create an empty raster template
template <- rast(extent = extent_nlcd, res = target_res, crs = crs(nlcd))

# Resample the original raster to match the new template
nlcd_resampled <- resample(nlcd, template, method = "near")  # Use "bilinear" for continuous data or "near" for categorical

# Convert raster to data frame for plotting 
nlcd_df <- as.data.frame(nlcd_resampled, na.rm = TRUE, xy = TRUE)

# Add land cover type labels
nlcd_df <- nlcd_df %>%
  mutate(LandType = case_when(
    Layer_1 == 11 ~ "Open Water",
    Layer_1 == 12 ~ "Perennial Ice/Snow",
    Layer_1 == 21 ~ "Developed, Open Space",
    Layer_1 == 22 ~ "Developed, Low Intensity",
    Layer_1 == 23 ~ "Developed, Medium Intensity",
    Layer_1 == 24 ~ "Developed, High Intensity",
    Layer_1 == 31 ~ "Barren Land (Rock/Sand/Clay)",
    Layer_1 == 41 ~ "Deciduous Forest",
    Layer_1 == 42 ~ "Evergreen Forest",
    Layer_1 == 43 ~ "Mixed Forest",
    Layer_1 == 51 ~ "Dwarf Scrub",
    Layer_1 == 52 ~ "Shrub/Scrub",
    Layer_1 == 71 ~ "Grassland/Herbaceous",
    Layer_1 == 72 ~ "Sedge/Herbaceous",
    Layer_1 == 73 ~ "Lichens",
    Layer_1 == 74 ~ "Moss",
    Layer_1 == 81 ~ "Pasture/Hay",
    Layer_1 == 82 ~ "Cultivated Crops",
    Layer_1 == 90 ~ "Woody Wetlands",
    Layer_1 == 95 ~ "Emergent Herbaceous Wetlands",
    TRUE ~ NA_character_  # Exclude unknown values
  )) %>%
  filter(!is.na(LandType))  # Remove rows with NA values

# Observe dataframe columns 
head(nlcd_df, 50)

# Initial data exploration: ensuring the nlcd data loads and displays properly 
# Create a heatmap with discrete categories in 'Layer_1'


ggplot(nlcd_df, aes(x = x, y = y, fill = LandType)) +
  geom_tile() +  # Create a heatmap-like plot with tiles
  scale_fill_viridis_d() +  # Use a discrete color scale (viridis is suitable for categorical data)
  theme_minimal() +  # Clean theme
  labs(title = "Land Cover Data",
       x = "Longitude",
       y = "Latitude",
       fill = "Land Cover Type") +
  coord_fixed()  # Fix aspect ratio to prevent distortion

```


# 3. Spatial Integration

```{r}
# Spatial Integration
  # Convert OBA bee observations to spatial points
  # Convert Oregon boundary to spatial object 


# Convert OBA data to spatial points
oba_sf <- st_as_sf(oba, coords = c("Long", "Lat"), crs = 4326)

# Transform to the same CRS as the NLCD raster (assuming the raster is in UTM)
oba_sf <- st_transform(oba_sf, crs = crs(nlcd_resampled))

# Convert Oregon boundary to spatial object 
or_boundary <- st_as_sf(or_boundary, coords = c("lon", "lat"), crs = crs(oba_sf))

# Plot the Oregon boundary
# Ensure that the Oregon boundary plot loads properly 
ggplot() +
  geom_sf(data = or_boundary, fill = "lightblue", color = "black") +
  theme_minimal() +
  labs(title = "Oregon Boundary",
       x = "Longitude",
       y = "Latitude")

```

# Crop oba and nlcd to or 
```{r}
# Clip OBA data to Oregon boundary
oba_sf <- st_crop(oba_sf, or_boundary)
str(oba_sf)

# Check the result
summary(oba_sf)
```

# Elevational Band Categorization
## Define elevation bands (Low, Middle, High).
## Assign OBA observations to corresponding elevation bands.

```{r}
# Divide data into elevational bands
elevation_bands <- cut(oba_sf$Elev, breaks = c(0, 500, 1000, 1500, Inf), 
                       labels = c("0-500m", "500-1000m", "1000-1500m", "1500m+"))

# Add to spatial data
oba_sf <- oba_sf %>%
  mutate(ElevationBand = elevation_bands) %>%
  filter(!is.na(ElevationBand))

head(oba_sf)
```
# Create initial visualizations and stats 
```{r}
# Manipulation and summarizing data in relevant ways (initial exploration of data)

summary(oba_sf)
colnames(oba_sf)

# Add counts for species
species_counts_location <- oba %>%
  group_by(Long, Lat) %>%
  summarise(Count = n())

# Scatter plot with point size by count
ggplot(species_counts_location, aes(x = Long, y = Lat, size = Count)) +
  geom_point(color = "blue", alpha = 0.6) +
  theme_minimal() +
  labs(title = "Species Count by Location",
       x = "Longitude",
       y = "Latitude",
       size = "Count")

library(ggplot2)
library(sf)

# Plot colored by ElevationBand
ggplot() +
  # Add Oregon boundary layer
  geom_sf(data = or_boundary, fill = "lightblue", color = "black", alpha = 0.5) +
  
  # OBA points colored by ElevationBand
  geom_sf(data = oba_sf, aes(color = ElevationBand), size = 0.5, alpha = 0.7) +

  theme_minimal() +
  labs(title = "OBA Occurrences Locations Colored by Elevation Bands",
       x = "Longitude",
       y = "Latitude",
       color = "Elevation Band") +
  theme(legend.position = "right")

```


# Creating buffers 
```{r oba-buffers}

# Change to crs that uses meters, not feet 
oba_sf <- st_transform(oba_sf, crs = 32610)

# Generate 500m buffers (one around each oba data point)
buffers_500m <- st_buffer(oba_sf, dist = 500)

# add buffer_id column to buffer
buffers_500m$buffer_id <- seq_len(nrow(buffers_500m))

#perform spatial join: this assigns each species point to a buffer by buffer_id
species_in_buffers <- st_join(oba_sf, buffers_500m, join = st_within)

```



## Calculating and organizing species richness for each buffer 
```{r vegan package richness}

community_matrix <- species_in_buffers %>%
    st_drop_geometry() %>%                    # Drop spatial data for tabular manipulation
    group_by(buffer_id, GenusSpec.x) %>%          # Replace `buffer_id` with your buffer identifier column
    summarize(count = n()) %>%                # Count occurrences
    pivot_wider(names_from = GenusSpec.x, values_from = count, values_fill = 0) %>%
    column_to_rownames("buffer_id")           # Set buffer IDs as row names

richness <- specnumber(community_matrix)
# richness$buffer_id <- as.integer(richness$buffer_id)  # Make buffer ids ints 
head(richness)

# Calculate total observations (sum of individuals per buffer)
total_observations <- rowSums(community_matrix)  # Sum across species for each buffer

# Calculate normalized richness (richness / total observations)
normalized_richness <- richness / total_observations

# Create a dataframe with buffer ID and normalized richness
richness_df <- tibble(
  buffer_id = as.integer(names(richness)),      # Convert names (buffer IDs) to integer
  normalized_richness = normalized_richness     # Add the normalized richness values
)

# Check result
head(richness_df)

# Join richness data back to the buffers
buffers_500m <- buffers_500m %>%
  left_join(richness_df, by = "buffer_id")

head(buffers_500m)

# Plot buffers with richness levels
ggplot() +
  geom_sf(data = buffers_500m, aes(fill = normalized_richness), color = "black", size = 0.1) +
  scale_fill_viridis_c(option = "plasma", na.value = "white") +
  labs(
    title = "Species Richness in Buffers (500m)",
    fill = "Richness",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal()
```

## Calculating species diversity (using Shannon's)
```{r}
## Calculating and organizing diversity (Shannon's index) for each buffer
# Using the vegan package to compute Shannon's diversity index

# Calculate Shannon's diversity index for each buffer
bee_diversity <- vegan::diversity(community_matrix, index = "shannon")

# Create a dataframe with buffer ID and Shannon's diversity
diversity_df <- tibble(
  buffer_id = as.integer(rownames(community_matrix)), # Extract buffer IDs as integers
  bee_diversity = bee_diversity               # Add diversity values
)

# Join diversity data back to the buffers
buffers_500m <- buffers_500m %>%
  left_join(diversity_df, by = "buffer_id")

# Check result
head(buffers_500m)
```

## Find landscape diversity for each buffer; append to buffers_500m
```{r}
buffers_500m <- st_transform(buffers_500m, crs = st_crs(nlcd_resampled))
# Spatial join to extract land cover types from NLCD raster within each buffer
land_cover_in_buffers <- extract(nlcd_resampled, buffers_500m, fun = function(x) {
  # Get the most frequent land cover type in the buffer area
  if (length(x) > 0) {
    # Count occurrences of each land cover type
    land_cover_count <- table(x)
    
    # Use vegan diversity() function to calculate Shannon's Div Index
    shannon_index <- vegan::diversity(as.numeric(land_cover_count), index = "shannon")
    
    return(shannon_index)
  } else {
    return(NULL)  # If no data within the buffer, return NULL
  }
})

# Rename cols to match buffers_500m
land_cover_in_buffers <- land_cover_in_buffers %>%
  rename(
    landscape_div = Layer_1,
    buffer_id = ID
  )

# Join diversity vals with buffers df 
# NOTE for testing: only run once 
buffers_500m <- buffers_500m %>%
  left_join(land_cover_in_buffers, by = "buffer_id")

head(buffers_500m)

```



```{r linear model-diversity}
bee_div <- buffers_500m$bee_diversity
names(bee_div) <- NULL
land_div <- buffers_500m$landscape_div

dats_div <- data.frame(bee_div = bee_div,
                       land_div=land_div)

mod <- lm(bee_div ~ scale(land_div), data=dats_div)
summary(mod)
```

```{r linear model- richness}
bee_rich <- buffers_500m$normalized_richness
names(bee_rich) <- NULL


dats_rich <- data.frame(bee_rich = bee_rich,
                       land_div=land_div)

mod2 <- lm(bee_rich ~ scale(land_div), data=dats_rich)
summary(mod2)


```

```{r}
# Set up plot colors
main_colors <- c("#008000", "#FF1493", "#1E90FF", "#FFA500")

# Scatter plot of landscape diversity vs. bee species richness
ggplot(buffers_500m, aes(x = landscape_div, y = normalized_richness)) +
  geom_point(aes(color = ElevationBand), alpha = 0.7) +  # Color by ElevationBand
  scale_color_manual(values = main_colors) +
  labs(
    title = "Landscape Diversity vs. Bee Species Richness",
    x = "Landscape Diversity (Shannon's Index)",
    y = "Normalized Bee Species Richness",
    caption = "The black line represents the linear regression fit."
  ) +
  geom_smooth(method = "lm", color = "black") +
  theme_minimal() +
  theme(
    plot.caption = element_text(hjust = 0)  # Left-align caption
  )


# Richness of 1: if a bee is all alone in a buffer (1/1)
# Better labels for elevations 
```

```{r}
# Scatter plot of landscape diversity vs. bee species richness
ggplot(buffers_500m, aes(x = landscape_div, y = bee_diversity)) +
  geom_point(aes(color = ElevationBand), alpha = 0.7) +  # Color by ElevationBand
  scale_color_manual(values = main_colors) + 
  labs(
    title = "Landscape Diversity vs. Bee Species Diversity",
    x = "Landscape Diversity (Shannon's Index)",
    y = "Bee Species Diversity (Shannon's Index)",
    caption = "The black line represents the linear regression fit."
  ) +
  geom_smooth(method = "lm", color = "black") +
  theme_minimal() +
  theme(
    plot.caption = element_text(hjust = 0)  # Left-align caption
  )

```

```{r}
# Plot landcover, oba, buffers 

# Import colors
library(RColorBrewer)
colors <- c(brewer.pal(9, "Dark2"), brewer.pal(9, "Paired"))

# Plot buffers around the bee observations
ggplot() +
  # Plot the Oregon boundary
  geom_sf(data = or_boundary, fill = NA, color = "black") +

  # Plot the land cover data (NLCD)
  geom_raster(data = nlcd_df, aes(x = x, y = y, fill = LandType)) +
  scale_fill_manual(values = colors, name = "Land Cover") + 

  # Plot the 500m buffers around OBA data
  geom_sf(data = buffers_500m, fill = "black", alpha = 1, color = "black") +

  # Customize the plot appearance
  labs(title = "Oregon Land Cover with Buffered Bee Observations",
       x = "Longitude",
       y = "Latitude", 
       caption = "Buffers (500m) are plotted as black circles")
  theme_minimal()
```

```{r}
# Create boxplots for diversity:

# Boxplot comparing landscape diversity for elevation bands 
boxplot(landscape_div~ElevationBand, data=buffers_500m, main="Landscape Diversity at Elevation Bands",
   xlab="Elevation (m)", ylab="Landscape Diversity",
   col = main_colors, border = "black", notch = TRUE, notchwidth = 0.5,
        medcol = "white", whiskcol = "black", boxwex = 0.5, outpch = 19,
        outcol = "black")

# Legend
legend("topright", legend = unique(buffers_500m$ElevationBand), 
       fill = main_colors, border = "black", title = "Elevation Bands", 
       cex = 0.6)  # cex to control font size


# Boxplot comparing bee diversity for elevation bands 
boxplot(bee_diversity~ElevationBand, data=buffers_500m, main="Bee Diversity at Elevation Bands",
   xlab="Elevation (m)", ylab="Bee Diversity", 
   col = main_colors, border = "black", notch = TRUE, notchwidth = 0.5,
        medcol = "white", whiskcol = "black", boxwex = 0.5, outpch = 19,
        outcol = "black")

# Legend
legend("bottomleft", legend = unique(buffers_500m$ElevationBand), 
       fill = main_colors, border = "black", title = "Elevation Bands", cex = 0.6)


# Boxplot comparing bee richness for elevation bands 
boxplot(bee_rich~ElevationBand, data=buffers_500m, main="Bee Richness at Elevation Bands",
   xlab="Elevation (m)", ylab="Bee Richness", 
   col = main_colors, border = "black", notch = TRUE, notchwidth = 0.5,
        medcol = "white", whiskcol = "black", boxwex = 0.5, outpch = 19,
        outcol = "black")

# Legend
legend("topright", legend = unique(buffers_500m$ElevationBand), 
       fill = main_colors, border = "black", title = "Elevation Bands", cex = 0.6)

```
# Summary stats for boxplots 
```{r}

# Summary stats for Landscape Diversity by Elevation Band
landscape_stats <- aggregate(landscape_div ~ ElevationBand, data = buffers_500m, 
                             FUN = function(x) c(mean = mean(x), median = median(x), 
                                                 min = min(x), max = max(x), 
                                                 IQR = IQR(x)))
# Unnest the results for better readability
landscape_stats <- do.call(data.frame, landscape_stats)

# Summary stats for Bee Diversity by Elevation Band
bee_diversity_stats <- aggregate(bee_diversity ~ ElevationBand, data = buffers_500m, 
                                  FUN = function(x) c(mean = mean(x), median = median(x), 
                                                      min = min(x), max = max(x), 
                                                      IQR = IQR(x)))
bee_diversity_stats <- do.call(data.frame, bee_diversity_stats)

# Summary stats for Bee Richness by Elevation Band
bee_richness_stats <- aggregate(bee_rich ~ ElevationBand, data = buffers_500m, 
                                 FUN = function(x) c(mean = mean(x), median = median(x), 
                                                     min = min(x), max = max(x), 
                                                     IQR = IQR(x)))
bee_richness_stats <- do.call(data.frame, bee_richness_stats)

# Print the results
print("Landscape Diversity Stats:")
landscape_stats

print("Bee Diversity Stats:")
bee_diversity_stats

print("Bee Richness Stats:")
bee_richness_stats

```

