---
title: "nlcd_oba"
author: "egb"
date: "2024-11-20"
output:
  pdf_document: default
  html_document: default
---
```{r load-libraries, echo=FALSE, results="hide", message=FALSE, warning=FALSE}
library(terra)
library(ggplot2)
library(dplyr)
library(sf)
library(tidyr)
```

NOTES from conversation with Lauren:  

- diversity of habitat types  
- don't subset to the two categories  

- calculate diversity at 100, 500, 1000 meters (changing buffer size) 
- a Buffer around every single bee (hypothetical place every bee would forage in) 
- find the pixel diversity in each of those buffers (to see if habitat heterogeneity improves bee diversity) 

- different generae associated with different habitat  

- in each buffer: number of species / number of observations (richness/diversity) 
      - richness factoring in variability in data collection 
      - relative richness to the number of observations  
      
- Split up the whole state / every bee observation into three elevational segments/bands  
- Look and bin some of the landcover data  
- Abundance-weight the pixels (shannon's, matrixes of diversity) 
- Use extract to create buffers  
- Don't do the label shuffling; use R lm() function  
- diversity begets diversity ecology (DBD) theory : paper that supports our research 


# Oregon basemap

```{r plot-or}

or_boundary <- read_sf(dsn = "data/Oregon_State_Boundary_-2975945127261102476/Oregon_State_Boundary.shp")
# str(or_boundary)

```




# 1. OBA Manipulation

```{r oba-import}

# Import OBA data:

oba <-
  read.csv("data/OBA_2018-2023_combined_dataset_working.csv")


```

```{r}
# Segment OBA data: 

# - Metadata we need:
#   - `Species`: The taxonomic identification of each bee (group of closely related species)
#   - `Genus`: The taxonomic identification of each bee (specific type within Genus)
#   - `Elevation`: The elevation (in meters) at which each bee was collected.
#   - `Coordinates`: Latitude (`Dec.Lat.`) and longitude (`Dec.Long.`) in decimal degrees.


# Subset oba data to our needed metadata
oba <- dplyr::select(oba, Genus, Species, Dec..Lat., Dec..Long., Elevation..m., MonthJul)
colnames(oba) <- c("Genus", "Spec", "Lat", "Long", "Elev", "Month")

# head(oba, 60)

# row count before cleaning: 195788
# nrow(oba)

# Make elevation values numeric 
oba$Elev <- as.numeric(oba$Elev)

# Filter data to just July 
# oba <- filter(oba, Month == "July")

```


```{r}
# Clean OBA data

# Remove elevation rows with null values 

for (col in colnames(oba)) {
  oba <- oba[oba[[col]] != "", ]
  oba <- oba[!is.na(oba[[col]]), ]
}
# Number of rows after cleaning: 41643
# nrow(oba)
# head(oba, 10)
```



```{r}
# Combine genus & species 
GenusSpec <- paste(oba$Genus, oba$Spec, sep=" ")

oba <- oba %>%
  mutate(GenusSpec)

# colnames(oba)
#nhead(oba, 10)
```


# 2. NLCD Manipulation

```{r nlcd-import, include=FALSE}
library(raster)

# Observe the dataset metadata here: https://www.arcgis.com/sharing/rest/content/items/9bbaa64718774bbfbf5c6ade0edf86d3/info/metadata/metadata.xml?format=default&output=html 

# Load landcover data as raster 
nlcd <- rast("data/NLCD_2016_Land_Cover_OR/NLCD_2016_Land_Cover_OR.img")



# LOWER RESOLUTION AND CREATE DF:
# Check original resolution
# Original: 98.4252 98.4252 
original_res <- res(nlcd)
# cat("Original resolution:", original_res, "\n")

# Create a new template raster with the desired resolution
# The larger the resolution, the easier to load (make this smaller for final run)
target_res <- 500  # Desired resolution in map units (e.g., meters)
extent_nlcd <- ext(nlcd)  # Get the extent of the original raster

# Create an empty raster template
template <- rast(extent = extent_nlcd, res = target_res, crs = crs(nlcd))

# Resample the original raster to match the new template
nlcd_resampled <- resample(nlcd, template, method = "near")  # Use "bilinear" for continuous data or "near" for categorical

#make sure it has the same CRS as the other data
 


print(nlcd_resampled)

# Convert raster to data frame for plotting 
nlcd_df <- as.data.frame(nlcd_resampled, na.rm = TRUE, xy = TRUE)

# Add land cover type labels
nlcd_df <- nlcd_df %>%
  mutate(LandType = case_when(
    Layer_1 == 11 ~ "Open Water",
    Layer_1 == 12 ~ "Perennial Ice/Snow",
    Layer_1 == 21 ~ "Developed, Open Space",
    Layer_1 == 22 ~ "Developed, Low Intensity",
    Layer_1 == 23 ~ "Developed, Medium Intensity",
    Layer_1 == 24 ~ "Developed, High Intensity",
    Layer_1 == 31 ~ "Barren Land (Rock/Sand/Clay)",
    Layer_1 == 41 ~ "Deciduous Forest",
    Layer_1 == 42 ~ "Evergreen Forest",
    Layer_1 == 43 ~ "Mixed Forest",
    Layer_1 == 51 ~ "Dwarf Scrub",
    Layer_1 == 52 ~ "Shrub/Scrub",
    Layer_1 == 71 ~ "Grassland/Herbaceous",
    Layer_1 == 72 ~ "Sedge/Herbaceous",
    Layer_1 == 73 ~ "Lichens",
    Layer_1 == 74 ~ "Moss",
    Layer_1 == 81 ~ "Pasture/Hay",
    Layer_1 == 82 ~ "Cultivated Crops",
    Layer_1 == 90 ~ "Woody Wetlands",
    Layer_1 == 95 ~ "Emergent Herbaceous Wetlands",
    TRUE ~ NA_character_  # Exclude unknown values
  )) %>%
  filter(!is.na(LandType))  # Remove rows with NA values

# Observe dataframe columns 
head(nlcd_df, 50)

# Create a heatmap with discrete categories in 'Layer_1'
ggplot(nlcd_df, aes(x = x, y = y, fill = LandType)) +
  geom_tile() +  # Create a heatmap-like plot with tiles
  scale_fill_viridis_d() +  # Use a discrete color scale (viridis is suitable for categorical data)
  theme_minimal() +  # Clean theme
  labs(title = "Land Cover Data",
       x = "Longitude",
       y = "Latitude",
       fill = "Land Cover Type") +
  coord_fixed()  # Fix aspect ratio to prevent distortion

```


# 3. Spational Integration
## Perform a spatial overlay to assign land cover types from NLCD to OBA points.

```{r}
# Spatial Integration
# Convert OBA bee observations to spatial points

# Convert OBA data to spatial points (assuming "Lat" and "Long" are in decimal degrees)
oba_sf <- st_as_sf(oba, coords = c("Long", "Lat"), crs = 4326)


# Transform to the same CRS as the NLCD raster (assuming the raster is in UTM)
oba_sf <- st_transform(oba_sf, crs = crs(nlcd_resampled))

# # 2. Plot the datasets
# ggplot() +
#   # Plot the land cover data (NLCD) as a raster
#   geom_raster(data = nlcd_df, aes(x = x, y = y, fill = LandType)) +
#   
#   # Overlay OBA bee observations as spatial points, colored by elevation
#   geom_sf(data = oba_sf, size = 1, aes(color = "red")) +
#   
#   # Customize the plot appearance
#   labs(title = "OBA Bee Observations Overlaid on NLCD Land Cover",
#        x = "Longitude",
#        y = "Latitude",
#        fill = "Land Cover Type",
#        color = "Elevation (m)") +
#   
#   theme_minimal() +
#   
#   # for spatial data handling
#   coord_sf()

```

# Crop oba and nlcd to or 
```{r}
# Make or crs match
# or_boundary <- st_transform(or_boundary, crs = st_crs(oba_sf))
# str(or_boundary)

# ISSUE: creates a set of points, not a continuous border 
# st_geometry_type(or_boundary_sf)

# Clip OBA data to Oregon boundary
# oba_cropped_sf <- st_crop(oba_sf, or_boundary_sf)
oba_sf <- st_crop(oba_sf, or_boundary)
str(oba_sf)

# Check the result
summary(oba_sf)

# Check the result
# summary(oba_cropped)
```
```{r}

or_boundary <- st_as_sf(or_boundary, coords = c("lon", "lat"), crs = crs(oba_sf))

# Test: plotting land cover and oba data over oregon border to ensure that oba data cropped to or border
# Plot using geom_sf for the Oregon boundary
# ggplot() +
#   # Plot the Oregon boundary
#   
#   # geom_sf(data = or_boundary, x = lat, y = lon, fill = NA, color = "black") +  # OR boundary
#   # geom_polygon(data = or_boundary, aes(x = lon, y = lat, group = group), fill = NA, color = "black")
#   geom_sf(data = or_boundary, fill = NA, color = "black") +
# 
#   # Plot the raster data (NLCD)
#   geom_raster(data = nlcd_df, aes(x = x, y = y, fill = LandType)) +
#   
#   # Plot the cropped OBA data (inside Oregon boundary)
#   geom_sf(data = oba_sf, aes(fill = "black"), size = .5) +
#   
#   # Customize the plot appearance
#   labs(title = "Cropped OBA Data, Oregon Boundary, and NLCD",
#        x = "Longitude",
#        y = "Latitude",
#        color = "OBA Observations") +
#   theme_minimal() 


```
# Elevational Band Categorization
## Define elevation bands (Low, Middle, High).
## Assign OBA observations to corresponding elevation bands.

```{r}

# crs(oba_sf)

# Divide data into elevational bands
elevation_bands <- cut(oba_sf$Elev, breaks = c(0, 500, 1000, 1500, Inf), 
                       labels = c("Low: 0-500m", "Medium: 500-1000m", "High: 1000-1500m", "Very High: 1500m-"))

# Add to spatial data
oba_sf <- oba_sf %>%
  mutate(ElevationBand = elevation_bands) %>%
  filter(!is.na(ElevationBand))

head(oba_sf)
```


# Creating buffers 
```{r oba-buffers}

# Idea: Put buffers around only bombus?
# Issue: creating buffers from same data we're counting on 
# Send Lauren Teams message about this issue

# Change to crs that uses meters, not feet 
oba_sf <- st_transform(oba_sf, crs = 32610)

# Generate 500m buffers (one around each oba data point)
buffers_500m <- st_buffer(oba_sf, dist = 500)

# add buffer_id column to buffer
buffers_500m$buffer_id <- seq_len(nrow(buffers_500m))

#perform spatial join: this assigns each species point to a buffer by buffer_id
species_in_buffers <- st_join(oba_sf, buffers_500m, join = st_within)

# Test 
# head(species_in_buffers)

# Try indexing oba data by oba buffer:
# school_robbery <- robberies_may[school_buffer,]
# oba_500m <- oba_sf[buffers_500m,]
# nrow(oba_500m)

# 
# # Plot buffers around the bee observations
# ggplot() +
#   # Plot the Oregon boundary
#   geom_sf(data = or_boundary, fill = NA, color = "black") +
# 
#   # Plot the land cover data (NLCD)
#   geom_raster(data = nlcd_df, aes(x = x, y = y, fill = LandType)) +
# 
#   # Plot the 100m buffers around OBA data
#   geom_sf(data = buffers_100m, fill = "blue", alpha = 0.3, color = "blue") +
# 
#   # Plot the 500m buffers around OBA data
#   geom_sf(data = buffers_500m, fill = "green", alpha = 0.3, color = "green") +
# 
#   # Plot the 1000m buffers around OBA data
#   geom_sf(data = buffers_1000m, fill = "red", alpha = 0.3, color = "red") +
# 
#   # Plot the OBA data points
#   geom_sf(data = oba_cropped_sf, aes(color = "black"), size = 0.1) +
# 
#   # Customize the plot appearance
#   labs(title = "Buffers Around OBA Bee Observations",
#        x = "Longitude",
#        y = "Latitude") +
#   theme_minimal()

```



## Calculating and organizing species richness for each buffer 
```{r vegan package richness}
# CHECKPOINT: successfully join richness data with buffers

# change to diversity? The 0 for 1 species 
# go through and make histograms of the actual data for sanity checks 
# could drop buffers with only 1 bee
# show for all months 
  # only map subset to July? 

# We could compare results for richness vs. diversity 
  # diversity() from vegan package - 
  # use shannons for both bee and landscape div 

library(vegan)
library(tibble) 

community_matrix <- species_in_buffers %>%
    st_drop_geometry() %>%                    # Drop spatial data for tabular manipulation
    group_by(buffer_id, GenusSpec.x) %>%          # Replace `buffer_id` with your buffer identifier column
    summarize(count = n()) %>%                # Count occurrences
    pivot_wider(names_from = GenusSpec.x, values_from = count, values_fill = 0) %>%
    column_to_rownames("buffer_id")           # Set buffer IDs as row names

richness <- specnumber(community_matrix)
# richness$buffer_id <- as.integer(richness$buffer_id)  # Make buffer ids ints 
head(richness)

# Calculate total observations (sum of individuals per buffer)
total_observations <- rowSums(community_matrix)  # Sum across species for each buffer

# Calculate normalized richness (richness / total observations)
normalized_richness <- richness / total_observations

# Create a dataframe with buffer ID and normalized richness
richness_df <- tibble(
  buffer_id = as.integer(names(richness)),      # Convert names (buffer IDs) to integer
  normalized_richness = normalized_richness     # Add the normalized richness values
)

# Check result
head(richness_df)

# Join richness data back to the buffers
buffers_500m <- buffers_500m %>%
  left_join(richness_df, by = "buffer_id")

head(buffers_500m)

# Plot buffers with richness levels
ggplot() +
  geom_sf(data = buffers_500m, aes(fill = normalized_richness), color = "black", size = 0.1) +
  scale_fill_viridis_c(option = "plasma", na.value = "white") +
  labs(
    title = "Species Richness in Buffers (500m)",
    fill = "Richness",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal()
```

## Calculating species diversity (using Shannon's)
```{r}
## Calculating and organizing diversity (Shannon's index) for each buffer
# Using the vegan package to compute Shannon's diversity index

# Calculate Shannon's diversity index for each buffer
bee_diversity <- vegan::diversity(community_matrix, index = "shannon")


# Create a dataframe with buffer ID and Shannon's diversity
diversity_df <- tibble(
  buffer_id = as.integer(rownames(community_matrix)), # Extract buffer IDs as integers
  bee_diversity = bee_diversity               # Add diversity values
)

# Join diversity data back to the buffers
buffers_500m <- buffers_500m %>%
  left_join(diversity_df, by = "buffer_id")

# Check result
head(buffers_500m)
```

## Find landscape diversity for each buffer; append to buffers_500m
```{r}
buffers_500m <- st_transform(buffers_500m, crs = st_crs(nlcd_resampled))
# Spatial join to extract land cover types from NLCD raster within each buffer
land_cover_in_buffers <- extract(nlcd_resampled, buffers_500m, fun = function(x) {
  # Get the most frequent land cover type in the buffer area
  if (length(x) > 0) {
    # Count occurrences of each land cover type
    land_cover_count <- table(x)
    
    # Use vegan diversity() function to calculate Shannon's Div Index
    shannon_index <- vegan::diversity(as.numeric(land_cover_count), index = "shannon")

    
    # Manual method: 
    # land_cover_freq <- land_cover_count / sum(land_cover_count)
    # shannon_index <- -sum(land_cover_freq * log(land_cover_freq), na.rm = TRUE)
    
    return(shannon_index)
  } else {
    return(NULL)  # If no data within the buffer, return NULL
  }
})

# Rename cols to match buffers_500m
land_cover_in_buffers <- land_cover_in_buffers %>%
  rename(
    landscape_div = Layer_1,
    buffer_id = ID
  )

# Check the result (it's a df)
# head(land_cover_in_buffers)

# Join diversity vals with buffers df 
# NOTE for testing: only run once 
buffers_500m <- buffers_500m %>%
  left_join(land_cover_in_buffers, by = "buffer_id")

head(buffers_500m)

```



```{r linear model-diversity}
bee_div <- buffers_500m$bee_diversity
names(bee_div) <- NULL
land_div <- buffers_500m$landscape_div

dats_div <- data.frame(bee_div = bee_div,
                       land_div=land_div)

mod <- lm(bee_div ~ scale(land_div), data=dats_div)
summary(mod)
```

```{r linear model- richness}
bee_rich <- buffers_500m$normalized_richness
names(bee_rich) <- NULL


dats_rich <- data.frame(bee_rich = bee_rich,
                       land_div=land_div)

mod2 <- lm(bee_rich ~ scale(land_div), data=dats_rich)
summary(mod2)


```

```{r}
# Set up plot colors
main_colors <- c("#008000", "#FF1493", "#1E90FF", "#FFA500")

# Scatter plot of landscape diversity vs. bee species richness
ggplot(buffers_500m, aes(x = landscape_div, y = normalized_richness)) +
  geom_point(aes(color = ElevationBand), alpha = 0.7) +  # Color by buffer_id
  scale_color_manual(values = main_colors) +
  labs(
    title = "Landscape Diversity vs. Bee Species Richness",
    x = "Landscape Diversity (Shannon Index)",
    y = "Normalized Bee Species Richness"
  ) +
  geom_smooth(method = "lm")+
  theme_minimal() 


# Richness of 1: if a bee is all alone in a buffer (1/1)
# Better labels for elevations 
```

```{r}
# Scatter plot of landscape diversity vs. bee species richness
ggplot(buffers_500m, aes(x = landscape_div, y = bee_diversity)) +
  geom_point(aes(color = ElevationBand), alpha = 0.7) +  # Color by buffer_id
  scale_color_manual(values = main_colors) + 
  labs(
    title = "Landscape Diversity vs. Bee Species Diversity",
    x = "Landscape Diversity (Shannon Index)",
    y = "Bee Species Diversity (Shannon Index)"
  ) +
  geom_smooth(method = "lm")+
  theme_minimal() 


# Richness of 1: if a bee is all alone in a buffer (1/1)
# Better labels for elevations 
```

```{r}
# Plot landcover, oba, buffers 

# Import colors
library(RColorBrewer)
colors <- c(brewer.pal(9, "Dark2"), brewer.pal(9, "Paired"))


# Plot buffers around the bee observations
ggplot() +
  # Plot the Oregon boundary
  geom_sf(data = or_boundary, fill = NA, color = "black") +

  # Plot the land cover data (NLCD)
  geom_raster(data = nlcd_df, aes(x = x, y = y, fill = LandType)) +
  scale_fill_manual(values = colors, name = "Land Cover") + 
  
  
  # Plot the OBA data points
  # geom_sf(data = oba_sf, aes(color = "black"), size = 0.1) +

  # Plot the 500m buffers around OBA data
  geom_sf(data = buffers_500m, fill = "black", alpha = 1, color = "black") +

  # Customize the plot appearance
  labs(title = "Buffers Around OBA Bee Observations",
       x = "Longitude",
       y = "Latitude", 
       caption = "Buffers (500m) are plotted as black circles")
  theme_minimal()
```

```{r}
# Create boxplots:

# Boxplot comparing landscape diversity for elevation bands 
boxplot(landscape_div~ElevationBand, data=buffers_500m, main="Landscape Diversity at Elevation Bands",
   xlab="Elevation (m)", ylab="Landscape Diversity",
   col = main_colors, border = "black", notch = TRUE, notchwidth = 0.5,
        medcol = "white", whiskcol = "black", boxwex = 0.5, outpch = 19,
        outcol = "black")

# Legend
legend("topright", legend = unique(buffers_500m$ElevationBand), 
       fill = main_colors, border = "black", title = "Elevation Bands", 
       cex = 0.6)  # cex to control font size


# Boxplot comparing bee diversity for elevation bands 
boxplot(bee_diversity~ElevationBand, data=buffers_500m, main="Bee Diversity at Elevation Bands",
   xlab="Elevation (m)", ylab="Bee Diversity", 
   col = main_colors, border = "black", notch = TRUE, notchwidth = 0.5,
        medcol = "white", whiskcol = "black", boxwex = 0.5, outpch = 19,
        outcol = "black")

# Legend
legend("bottomleft", legend = unique(buffers_500m$ElevationBand), 
       fill = main_colors, border = "black", title = "Elevation Bands", cex = 0.6)

```

