---
title: "Exploring the Relationship Between Landscape Diversity and both Bee Diversity and Normalized Richess"
author: "Eliza Black and Emmie Sullivan"
date: "Dec. 2024"
output:
  pdf_document:
    latex_engine: xelatex
  html_document: default
---

# Introduction:

## Relevant Background and References:

Wild and managed pollinators benefit society in various ways, including bolstering food security, supporting livelihoods, enriching cultural values, and sustaining biodiversity and ecosystem health (Potts et al., 2016). This has made the study of pollinators a prominent subject in ecology. Pollinator populations are experiencing significant declines due to habitat loss and land-use changes, such as deforestation and agricultural expansion, as well as wider "threats to global food supply and the stability of wider pollination systems in non-crop vegetation... [and] disruptions of numerous plant-pollinator systems” (Mader et al., 2011). 

It has been established that "diversity begets diversity", that is, local environmental heterogeneity is a key factor driving community richness (Rosenzweig, 1995, as cited in Ponisio, 2020). Previous research has shown that local fire severity is a strong predictor of wild bee diversity in mixed-conifer forests in southwestern Oregon (Galbraith et al, 2019), but few studies have been conducted state-wide to examine the relationship between bee diversity and landscape diversity overall. 

This study tests how landscape diversity impacts bee diversity and richness. Given the pressing threats to bee populations and their critical ecological roles, it is vital to explore how landscape diversity supports pollinator diversity. Understanding this relationship provides insight into habitat conditions that support bee populations. Identifying the habitat diversity levels that best support bee diversity can inform land management strategies to protect these populations.

# Hypotheses:

## Hypothesis 1

- Null Hypothesis (H0): There is no statistically significant relationship between landscape diversity and bee species diversity (measured by the Shannon-Wiener index).
- Alternative Hypothesis (H1): There is a statistically significant positive relationship between landscape diversity and bee species diversity (measured by the Shannon-Wiener index).

## Hypothesis 2

- Null Hypothesis (H0): There is no statistically significant relationship between landscape diversity and normalized bee species richness.
- Alternative Hypothesis (H1): There is a statistically significant positive relationship between landscape diversity and normalized bee species richness.

## Hypothesis 3

- Null Hypothesis (H0): There is no difference in species richness and diversity between elevation bands (0-500m, 500-1000m, 1000-1500m, 1500m+).
- Alternative Hypothesis (H1): The relationship between landscape diversity and both normalized bee species richness and diversity will vary significantly across different elevational bands. Specifically, at higher elevations, landscape diversity may have a weaker effect on bee species richness and diversity due to environmental constraints such as reduced floral resources and harsher weather. Conversely, the effect of landscape diversity on bee diversity and richness will be more pronounced at lower elevations. 


## Data Sources:

1. **Oregon Bee Atlas (OBA)**  
   - **Description**: Dataset compiled by citizen scientists and professional entomologists throughout Oregon. Includes bee species, geographic, and environmental metadata.
   - **Attributes**:
     - Species: Taxonomic identification of each bee recorded.
     - Elevation: Elevation (in meters) at which each bee was collected.
     - Coordinates: Latitude and longitude in decimal degrees.

2. **National Land Cover Database (NLCD) - Oregon**  
   - **Description**: Comprehensive land cover classifications, clipped to Oregon’s state boundary. Projected in Oregon Lambert (EPSG:2992).
   - **Attributes**:
     - 16 land cover classifications including grassland, forested areas, and developed land.
   - **Publication Details**:
     - Published: January 2019
     - Updated: November 2023
     - Publisher: U.S. Geological Survey (USGS)

3. **Oregon State Boundary Dataset**  
   - **Description**: Jurisdictional and cartographic perimeters of Oregon. Useful for resource management and mapping.


# Load in necessary libraries:
```{r load-libraries, echo=FALSE, results="hide", message=FALSE, warning=FALSE}
library(terra)
library(ggplot2)
library(dplyr)
library(sf)
library(tidyr)
library(vegan)
library(tibble) 
```

# Load in Oregon basemap:
```{r plot-or, warning=FALSE, message=FALSE}
or_boundary <- read_sf(dsn = "data/Oregon_State_Boundary_-2975945127261102476/Oregon_State_Boundary.shp")
# str(or_boundary)
```

# Load in OBA Data:
```{r oba-import, warning=FALSE, message=FALSE}
oba <-
  read.csv("data/OBA_2018-2023_combined_dataset_working.csv")
```

# Segment OBA data:
```{r oba-segment, warning=FALSE, message=FALSE}

# - Metadata we need:
#   - `Species`: The taxonomic identification of each bee (group of closely related species)
#   - `Genus`: The taxonomic identification of each bee (specific type within Genus)
#   - `Elevation`: The elevation (in meters) at which each bee was collected.
#   - `Coordinates`: Latitude (`Dec.Lat.`) and longitude (`Dec.Long.`) in decimal degrees.

# Subset oba data to our needed metadata
oba <- dplyr::select(oba, Genus, Species, Dec..Lat., Dec..Long., Elevation..m., MonthJul)
colnames(oba) <- c("Genus", "Spec", "Lat", "Long", "Elev", "Month")

# Make elevation values numeric 
oba$Elev <- as.numeric(oba$Elev)

```

# Clean OBA data:
```{r oba-clean, warning=FALSE, message=FALSE}
# Remove elevation rows with null values 

for (col in colnames(oba)) {
  oba <- oba[oba[[col]] != "", ]
  oba <- oba[!is.na(oba[[col]]), ]
}
# Number of rows after cleaning: 41643
# nrow(oba)
```

# Combine genus & species columns in OBA data:
```{r genus-spec, warning=FALSE, message=FALSE}
# Combine genus & species 
GenusSpec <- paste(oba$Genus, oba$Spec, sep=" ")

oba <- oba %>%
  mutate(GenusSpec)

```

# Load in NLCD - Oregon (National Land Cover Database):
```{r nlcd-import, include=FALSE, warning=FALSE, message=FALSE}
library(raster)

# Observe the dataset metadata here: https://www.arcgis.com/sharing/rest/content/items/9bbaa64718774bbfbf5c6ade0edf86d3/info/metadata/metadata.xml?format=default&output=html 

# Load landcover data as raster 
nlcd <- rast("data/NLCD_2016_Land_Cover_OR/NLCD_2016_Land_Cover_OR.img")

# Check original resolution
# Original: 98.4252 98.4252 ft
# res(nlcd)

```

# NLCD Manipulation:
```{r nlcd-manip, warning=FALSE, message=FALSE}
# Create a new template raster with the desired resolution
# The larger the resolution, the easier to load

# NOTE: need to reproject the raster before changing the crs to use meters as the unit due to the large file size 
target_res <- 500  # Desired resolution in original map units (feet)
extent_nlcd <- ext(nlcd)  # Get the extent of the original raster

# Create an empty raster template
template <- rast(extent = extent_nlcd, res = target_res, crs = crs(nlcd))

# Resample the original raster to match the new template
nlcd_resampled <- resample(nlcd, template, method = "near")  # Use "bilinear" for continuous data or "near" for categorical

# Convert raster to data frame for plotting 
nlcd_df <- as.data.frame(nlcd_resampled, na.rm = TRUE, xy = TRUE)

# Add land cover type labels
nlcd_df <- nlcd_df %>%
  mutate(LandType = case_when(
    Layer_1 == 11 ~ "Open Water",
    Layer_1 == 12 ~ "Perennial Ice/Snow",
    Layer_1 == 21 ~ "Developed, Open Space",
    Layer_1 == 22 ~ "Developed, Low Intensity",
    Layer_1 == 23 ~ "Developed, Medium Intensity",
    Layer_1 == 24 ~ "Developed, High Intensity",
    Layer_1 == 31 ~ "Barren Land (Rock/Sand/Clay)",
    Layer_1 == 41 ~ "Deciduous Forest",
    Layer_1 == 42 ~ "Evergreen Forest",
    Layer_1 == 43 ~ "Mixed Forest",
    Layer_1 == 51 ~ "Dwarf Scrub",
    Layer_1 == 52 ~ "Shrub/Scrub",
    Layer_1 == 71 ~ "Grassland/Herbaceous",
    Layer_1 == 72 ~ "Sedge/Herbaceous",
    Layer_1 == 73 ~ "Lichens",
    Layer_1 == 74 ~ "Moss",
    Layer_1 == 81 ~ "Pasture/Hay",
    Layer_1 == 82 ~ "Cultivated Crops",
    Layer_1 == 90 ~ "Woody Wetlands",
    Layer_1 == 95 ~ "Emergent Herbaceous Wetlands",
    TRUE ~ NA_character_  # Exclude unknown values
  )) %>%
  filter(!is.na(LandType))  # Remove rows with NA values

# Observe dataframe columns 
head(nlcd_df)

# Initial data exploration: ensuring the nlcd data loads and displays properly 
ggplot(nlcd_df, aes(x = x, y = y, fill = LandType)) +
  geom_tile() +  # Create a heatmap-like plot with tiles
  scale_fill_viridis_d() +  # Use a discrete color scale (viridis is suitable for categorical data)
  theme_minimal() +  # Clean theme
  labs(title = "Land Cover Data",
       x = "Longitude",
       y = "Latitude",
       fill = "Land Cover Type") +
  coord_fixed()  # Fix aspect ratio to prevent distortion

```


# 3. Spatial Integration:
```{r spat-integration, warning=FALSE, message=FALSE}
# Spatial Integration
  # Convert OBA bee observations to spatial points
  # Convert Oregon boundary to spatial object 


# Convert OBA data to spatial points
oba_sf <- st_as_sf(oba, coords = c("Long", "Lat"), crs = 4326)

# Transform to the same CRS as the NLCD raster (assuming the raster is in UTM)
oba_sf <- st_transform(oba_sf, crs = crs(nlcd_resampled))

# Convert Oregon boundary to spatial object 
or_boundary <- st_as_sf(or_boundary, coords = c("lon", "lat"), crs = crs(oba_sf))

# Plot the Oregon boundary
# Ensure that the Oregon boundary plot loads properly 
ggplot() +
  geom_sf(data = or_boundary, fill = "lightblue", color = "black") +
  theme_minimal() +
  labs(title = "Oregon Boundary",
       x = "Longitude",
       y = "Latitude")

```

# Crop oba and nlcd to or:
```{r crop-oba, warning=FALSE, message=FALSE}
# Clip OBA data to Oregon boundary
oba_sf <- st_crop(oba_sf, or_boundary)
str(oba_sf)

```

# Elevational Band Categorization
## Define elevation bands (Low, Middle, High).
## Assign OBA observations to corresponding elevation bands.

```{r elev-bands, warning=FALSE, message=FALSE}
# Divide data into elevational bands
elevation_bands <- cut(oba_sf$Elev, breaks = c(0, 500, 1000, 1500, Inf), 
                       labels = c("0-500m", "500-1000m", "1000-1500m", "1500m+"))

# Add to spatial data
oba_sf <- oba_sf %>%
  mutate(ElevationBand = elevation_bands) %>%
  filter(!is.na(ElevationBand))

head(oba_sf)
```

# Create initial visualizations and stats:
```{r initial-visuals, warning=FALSE, message=FALSE}
# Manipulation and summarizing data in relevant ways (initial exploration of data)

# Print summary stats for oba_sf
summary(oba_sf)

# Add counts for species
species_counts_location <- oba %>%
  group_by(Long, Lat) %>%
  summarise(Count = n())

# Scatter plot with point size by count
ggplot(species_counts_location, aes(x = Long, y = Lat, size = Count)) +
  geom_point(color = "blue", alpha = 0.6) +
  theme_minimal() +
  labs(title = "Species Count by Location",
       x = "Longitude",
       y = "Latitude",
       size = "Count")

# Plot colored by ElevationBand
ggplot() +
  # Add Oregon boundary layer
  geom_sf(data = or_boundary, fill = "lightblue", color = "black", alpha = 0.5) +
  
  # OBA points colored by ElevationBand
  geom_sf(data = oba_sf, aes(color = ElevationBand), size = 0.5, alpha = 0.7) +

  theme_minimal() +
  labs(title = "OBA Occurrences Locations Colored by Elevation Bands",
       x = "Longitude",
       y = "Latitude",
       color = "Elevation Band") +
  theme(legend.position = "right")

```

# Creating buffers:
```{r oba-buffers, warning=FALSE, message=FALSE}
# Change to crs that uses meters, not feet 
oba_sf <- st_transform(oba_sf, crs = 32610)

# Generate 500m buffers (one around each oba data point)
buffers_500m <- st_buffer(oba_sf, dist = 500)

# add buffer_id column to buffer
buffers_500m$buffer_id <- seq_len(nrow(buffers_500m))

#perform spatial join: this assigns each species point to a buffer by buffer_id
species_in_buffers <- st_join(oba_sf, buffers_500m, join = st_within)
```

## Calculating and organizing normalized species richness for each buffer:
```{r oba-richess, warning=FALSE, message=FALSE}

community_matrix <- species_in_buffers %>%
    st_drop_geometry() %>%                    
    group_by(buffer_id, GenusSpec.x) %>%         
    summarize(count = n()) %>%                # Count occurrences
    pivot_wider(names_from = GenusSpec.x, values_from = count, values_fill = 0) %>%
    column_to_rownames("buffer_id")           # Set buffer IDs as row names

richness <- specnumber(community_matrix)
# richness$buffer_id <- as.integer(richness$buffer_id)  # Make buffer ids ints 
head(richness)

# Calculate total observations (sum of individuals per buffer)
total_observations <- rowSums(community_matrix)  # Sum across species for each buffer

# Calculate normalized richness (richness / total observations)
normalized_richness <- richness / total_observations

# Create a dataframe with buffer ID and normalized richness
richness_df <- tibble(
  buffer_id = as.integer(names(richness)),      # Convert names (buffer IDs) to integer
  normalized_richness = normalized_richness     # Add the normalized richness values
)

# Check result
head(richness_df)

# Join richness data back to the buffers
buffers_500m <- buffers_500m %>%
  left_join(richness_df, by = "buffer_id")

head(buffers_500m)

# Plot buffers with richness levels
ggplot() +
  geom_sf(data = buffers_500m, aes(fill = normalized_richness), color = "black", size = 0.1) +
  scale_fill_viridis_c(option = "plasma", na.value = "white") +
  labs(
    title = "Species Richness in Buffers (500m)",
    fill = "Richness",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal()
```

## Calculating species diversity (using Shannon's):
```{r oba-diversity, warning=FALSE, message=FALSE}
## Calculating and organizing diversity (Shannon's index) for each buffer
# Using the vegan package to compute Shannon's diversity index

# Calculate Shannon's diversity index for each buffer
bee_diversity <- vegan::diversity(community_matrix, index = "shannon")

# Create a dataframe with buffer ID and Shannon's diversity
diversity_df <- tibble(
  buffer_id = as.integer(rownames(community_matrix)), # Extract buffer IDs as integers
  bee_diversity = bee_diversity               # Add diversity values
)

# Join diversity data back to the buffers
buffers_500m <- buffers_500m %>%
  left_join(diversity_df, by = "buffer_id")

# Check result
head(buffers_500m)
```

## Find landscape diversity for each buffer; append to buffers_500m:
```{r}
buffers_500m <- st_transform(buffers_500m, crs = st_crs(nlcd_resampled))
# Spatial join to extract land cover types from NLCD raster within each buffer
land_cover_in_buffers <- extract(nlcd_resampled, buffers_500m, fun = function(x) {
  # Get the most frequent land cover type in the buffer area
  if (length(x) > 0) {
    # Count occurrences of each land cover type
    land_cover_count <- table(x)
    
    # Use vegan diversity() function to calculate Shannon's Div Index
    shannon_index <- vegan::diversity(as.numeric(land_cover_count), index = "shannon")
    
    return(shannon_index)
  } else {
    return(NULL)  # If no data within the buffer, return NULL
  }
})

# Rename cols to match buffers_500m
land_cover_in_buffers <- land_cover_in_buffers %>%
  rename(
    landscape_div = Layer_1,
    buffer_id = ID
  )

# Join diversity vals with buffers df 
# NOTE for testing: only run once 
buffers_500m <- buffers_500m %>%
  left_join(land_cover_in_buffers, by = "buffer_id")

head(buffers_500m)

```

# Hypothesis Tests

## Hypotheses 1 and 2

- **Linear model tests**
    - The purpose is to estimate the strength and direction of the realtionshpi between lanscape diversity and bee species richness and diversity.
    - **Resulting statistics:**
        - **Coefficient:** Shows how each unit change in predictors (land cover) affects species richness or species diversity.
        - **p-value:** Shows the significance of land cover on species richness or diversity.
        - **R-squared:** Indicates how much variation in species richness or diversity is explained by land cover.

## Hypothesis 3: Species richness differs significantly between elevation bands

- **ANOVA**
    - **Analysis of Variance (ANOVA):**
        - A statistical formula that compares variances across the means of different groups.
        - It is used in the study to compare landscape diversity, bee diversity, and bee richness values across the different elevation bands.
    - **Resulting statistics:**
        - **Mean Square (Mean Sq):** An estimate of the variance associated with each factor.
        - **p-value:** Tests the null hypothesis that the factor has no effect (i.e., the means are equal across groups). Very low p-values (< 0.001) suggest that the factor or interaction significantly affects the response variable.
        - **Residuals:** Illustrate unexplained variance.


# Create statistics for oba diversity and richness linear models 
```{r lm-stats, warning=FALSE, message=FALSE}

# Reusable function for creating linear models and summarizing stats
linear_model <- function(dependent_var, independent_var, dataset, scale_independent = TRUE) {
  # Ensure dependent var is not named
  names(dependent_var) <- NULL
  
  # Create data frame for the model
  data <- data.frame(dependent_var = dependent_var, independent_var = independent_var)
  
  # Scale independent var if needed
  if (scale_independent) {
    model <- lm(dependent_var ~ scale(independent_var), data = data)
  } else {
    model <- lm(dependent_var ~ independent_var, data = data)
  }
  
  # Return the model summary
  return(summary(model))
}

bee_div <- buffers_500m$bee_diversity
bee_rich <- buffers_500m$normalized_richness
land_div <- buffers_500m$landscape_div

# Model for bee diversity
summary_div <- linear_model(dependent_var = bee_div, independent_var = land_div, dataset = buffers_500m)

# Model for normalized bee richness
summary_rich <- linear_model(dependent_var = bee_rich, independent_var = land_div, dataset = buffers_500m)

# Print summaries
summary_div
summary_rich

```
## Results:
Both linear models show statistically significant relationships between bee richness/diversity and the scaled landscape diversity, with p-values less than 2e-16. However, the very low R-squared values (0.0155 and 0.0244), indicate that the independent variable explains only a small amount of the variability in the bee richness/diversity.


# Plot linear models: Landscape diversity vs. bee richness / bee diversity 
```{r lm-plots, warning=FALSE, message=FALSE}

# Set up plot colors
main_colors <- c("#008000", "#FF1493", "#1E90FF", "#FFA500")

# Scatter plot of landscape diversity vs. bee species richness
ggplot(buffers_500m, aes(x = landscape_div, y = normalized_richness)) +
  geom_point(aes(color = ElevationBand), alpha = 0.7) +  # Color by ElevationBand
  scale_color_manual(values = main_colors) +
  geom_smooth(method = "lm", color = "black") +
  labs(
    title = "Landscape Diversity vs. Bee Species Richness",
    x = "Landscape Diversity (Shannon's Index)",
    y = "Normalized Bee Species Richness",
    caption = "The black line represents the linear regression fit."
  ) +
  theme_minimal() +
  theme(
    plot.caption = element_text(hjust = 0)  # Left-align caption
  )

# Scatter plot of landscape diversity vs. bee species richness
ggplot(buffers_500m, aes(x = landscape_div, y = bee_diversity)) +
  geom_point(aes(color = ElevationBand), alpha = 0.7) +  # Color by ElevationBand
  scale_color_manual(values = main_colors) + 
  geom_smooth(method = "lm", color = "black") +
  labs(
    title = "Landscape Diversity vs. Bee Species Diversity",
    x = "Landscape Diversity (Shannon's Index)",
    y = "Bee Species Diversity (Shannon's Index)",
    caption = "The black line represents the linear regression fit.") + 
  theme_minimal() +
  theme(
    plot.caption = element_text(hjust = 0)  # Left-align caption
  )
```

# Create map to display buffers overlaying nlcd
```{r buffer-plot, warning=FALSE, message=FALSE}
# Plot landcover, oba, buffers 

# Import colors
library(RColorBrewer)
colors <- c(brewer.pal(9, "Dark2"), brewer.pal(9, "Paired"))

# Plot buffers around the bee observations
ggplot() +
  # Plot the Oregon boundary
  geom_sf(data = or_boundary, fill = NA, color = "black") +

  # Plot the land cover data (NLCD)
  geom_raster(data = nlcd_df, aes(x = x, y = y, fill = LandType)) +
  scale_fill_manual(values = colors, name = "Land Cover") + 

  # Plot the 500m buffers around OBA data
  geom_sf(data = buffers_500m, fill = "black", alpha = 1, color = "black") +

  # Customize the plot appearance
  labs(title = "Oregon Land Cover with Buffered Bee Observations",
       x = "Longitude",
       y = "Latitude", 
       caption = "Buffers (500m) are plotted as black circles") + 
  theme_minimal()
```

# Exploring secondary relationships: 
# Create boxplots to view relationship between elevation bands and other variables
```{r elev-boxplots, warning=FALSE, message=FALSE}
# Create boxplots for diversity:

# Boxplot comparing landscape diversity for elevation bands 
boxplot(landscape_div~ElevationBand, data=buffers_500m, main="Landscape Diversity at Elevation Bands",
   xlab="Elevation (m)", ylab="Landscape Diversity",
   col = main_colors, border = "black", notch = TRUE, notchwidth = 0.5,
        medcol = "white", whiskcol = "black", boxwex = 0.5, outpch = 19,
        outcol = "black")

# Legend
legend("topright", legend = unique(buffers_500m$ElevationBand), 
       fill = main_colors, border = "black", title = "Elevation Bands", 
       cex = 0.6)  # cex to control font size


# Boxplot comparing bee diversity for elevation bands 
boxplot(bee_diversity~ElevationBand, data=buffers_500m, main="Bee Diversity at Elevation Bands",
   xlab="Elevation (m)", ylab="Bee Diversity", 
   col = main_colors, border = "black", notch = TRUE, notchwidth = 0.5,
        medcol = "white", whiskcol = "black", boxwex = 0.5, outpch = 19,
        outcol = "black")

# Legend
legend("bottomleft", legend = unique(buffers_500m$ElevationBand), 
       fill = main_colors, border = "black", title = "Elevation Bands", cex = 0.6)


# Boxplot comparing bee richness for elevation bands 
boxplot(bee_rich~ElevationBand, data=buffers_500m, main="Bee Richness at Elevation Bands",
   xlab="Elevation (m)", ylab="Bee Richness", 
   col = main_colors, border = "black", notch = TRUE, notchwidth = 0.5,
        medcol = "white", whiskcol = "black", boxwex = 0.5, outpch = 19,
        outcol = "black")

# Legend
legend("topright", legend = unique(buffers_500m$ElevationBand), 
       fill = main_colors, border = "black", title = "Elevation Bands", cex = 0.6)

```

# Summary stats for boxplots:
```{r}

# Summary stats for Landscape Diversity by Elevation Band
landscape_stats <- aggregate(landscape_div ~ ElevationBand, data = buffers_500m, 
                             FUN = function(x) c(mean = mean(x), median = median(x), 
                                                 min = min(x), max = max(x), 
                                                 IQR = IQR(x)))
# Unnest the results for better readability
landscape_stats <- do.call(data.frame, landscape_stats)

# Summary stats for Bee Diversity by Elevation Band
bee_diversity_stats <- aggregate(bee_diversity ~ ElevationBand, data = buffers_500m, 
                                  FUN = function(x) c(mean = mean(x), median = median(x), 
                                                      min = min(x), max = max(x), 
                                                      IQR = IQR(x)))
bee_diversity_stats <- do.call(data.frame, bee_diversity_stats)

# Summary stats for Bee Richness by Elevation Band
bee_richness_stats <- aggregate(bee_rich ~ ElevationBand, data = buffers_500m, 
                                 FUN = function(x) c(mean = mean(x), median = median(x), 
                                                     min = min(x), max = max(x), 
                                                     IQR = IQR(x)))
bee_richness_stats <- do.call(data.frame, bee_richness_stats)

# Print the results
landscape_stats
bee_diversity_stats
bee_richness_stats

```

## Results: 
- Highest bee diversity at lowest elevation band
- Landscape diversity decreases as elevation increases
- Lowest bee richness at lowest elevation band

# Test the hypothesis that the relationship between landscape diversity and bee species richness and diversity varies significantly across different elevational bands:
```{r}

# Summarize species richness and diversity by elevation band
summarized_by_elevation <- buffers_500m %>%
  group_by(ElevationBand) %>%
  summarise(
    mean_richness = mean(normalized_richness, na.rm = TRUE),  # Mean of normalized richness
    median_richness = median(normalized_richness, na.rm = TRUE),  # Median of normalized richness
    mean_diversity = mean(bee_diversity, na.rm = TRUE),  # Mean of bee diversity
    median_diversity = median(bee_diversity, na.rm = TRUE),  # Median of bee diversity
    # Optionally add other summary statistics like SD, count, etc.
    sd_richness = sd(normalized_richness, na.rm = TRUE),  # Standard deviation of richness
    sd_diversity = sd(bee_diversity, na.rm = TRUE)  # Standard deviation of diversity
  )

# View the summarized data
print(summarized_by_elevation)

# ANOVA for normalized richness across elevation bands
anova_richness <- aov(normalized_richness ~ landscape_div * ElevationBand, data = buffers_500m)
summary(anova_richness)

# ANOVA for bee diversity across elevation bands
anova_diversity <- aov(bee_diversity ~ landscape_div * ElevationBand, data = buffers_500m)
summary(anova_diversity)

# The ANOVA results indicate that the relationship between landscape diversity and bee richness as well as bee diversity  significantly varies across different elevation bands.

# Facet grid plot for normalized richness
ggplot(buffers_500m, aes(x = landscape_div, y = normalized_richness)) +
  geom_point(aes(color = ElevationBand)) +
  geom_smooth(method = "lm", aes(group = ElevationBand), se = FALSE) +
  facet_wrap(~ ElevationBand) +
  labs(title = "Normalized Richness by Landscape Diversity and Elevation Band",
       x = "Landscape Diversity", y = "Normalized Richness") +
  theme_minimal()

# Facet grid plot for bee diversity
ggplot(buffers_500m, aes(x = landscape_div, y = bee_diversity)) +
  geom_point(aes(color = ElevationBand)) +
  geom_smooth(method = "lm", aes(group = ElevationBand), se = FALSE) +
  facet_wrap(~ ElevationBand) +
  labs(title = "Bee Diversity by Landscape Diversity and Elevation Band",
       x = "Landscape Diversity", y = "Bee Diversity") +
  theme_minimal()

```

# Conclusion: summarize findings and address orignal hypotheses

## Landscape Diversity in Relation to Bee Diversity

There is a statistically significant positive relationship between bee diversity and landscape diversity. However, the effect size is small (low \(R^2\)), meaning landscape diversity explains very little of the variation in bee diversity.

## Landscape Diversity in Relation to Bee Richness

There is a statistically significant negative relationship between bee richness and landscape diversity. However, the effect size is small (low \(R^2\)), meaning landscape diversity explains very little of the variation in bee richness.

## Comparisons

- **Effect Size**: The slope is stronger for bee diversity than bee richness (absolute value of coefficients: \(0.0969\) vs \(0.0339\)).
- **Model Fit**: The bee richness model explains slightly more variability (\(R^2 = 0.02442\)) than the bee diversity model (\(R^2 = 0.01552\)).
- **Residual Standard Error**: The errors are smaller in the bee richness model (\(0.2145\)) compared to bee diversity (\(0.772\)).

## Variance by Elevation of Landscape Diversity’s Relation to Bee Diversity and Bee Richness

Landscape diversity and elevation band both significantly predict the response variable (bee diversity or richness), as indicated by the low p-values (< \(2 \times 10^{-16}\)) and large F-values.

The significant interaction between these two predictors (landscape diversity and elevation band) suggests that the relationship between landscape diversity and the response variables (bee richness and diversity) varies across elevation bands.

The high residual variance, indicated by the residual sum of squares, illustrates that a large portion of the variance is not explained by the model and that other variables are influencing the response variables.

## Secondary Relationships

- **Mean Landscape Diversity**: Mean landscape diversity decreases as elevation increases, with the highest values in the \(0-500\)m band (mean = \(1.27\)) and the lowest in the \(1500\)m+ band (mean = \(0.64\)).
- **Bee Diversity**: Higher bee diversity is observed at low and high elevations (mean = \(2.06\) at \(0-500\)m and \(1.79\) at \(1500\)m+).
- **Bee Richness**: Bee richness peaks at mid-elevations, with the highest mean at \(1000-1500\)m (mean = \(0.47\)), rising from \(0.35\) at \(0-500\)m and slightly declining to \(0.46\) at \(1500\)m+.

## Limits and Considerations

Data collected by citizen scientists introduces potential sampling biases. For example, the large number of buffers housing a single data point could reflect instances where scientists only recorded specialized bees. The study compares bee diversity and richness to clarify these potential biases and observe how they manifest differently.

## Recommendations for Future Studies

A future modification could involve grouping land cover types differently, such as combining developed land categories. This could reduce sampling bias and enhance the model’s explanatory power.
\newpage

## References

Galbraith, S. M., J. H. Cane, A. R. Moldenke, and J. W. Rivers. 2019. Wild bee diversity increases with local fire
severity in a fire-prone landscape. Ecosphere 10(4):e02668. 10.1002/ecs2.2668

New, T.R.Eric Mader, et al: attracting native pollinators. Protecting North America’s bees and butterflies. The Xerces Society Guide. J Insect Conserv 15, 611–612 (2011). https://doi.org/10.1007/s10841-011-9409-4

Potts, S., Imperatriz-Fonseca, V., Ngo, H. et al. Safeguarding pollinators and their values to human well-being. Nature 540, 220–229 (2016). https://doi.org/10.1038/nature20588

Ponisio LC. Pyrodiversity promotes interaction complementarity and population resistance. Ecol Evol. 2020 Mar 26;10(10):4431-4447. doi: 10.1002/ece3.6210. PMID: 32489608; PMCID: PMC7246207.

Rosenzweig, M. L. (1995). Species diversity in space and time. Cambridge, UK: Cambridge University Press.











